---
title: "<center>Assignment4</center>"
author: "<center>Dev Amin, Manan Patel & Prisha Shah</center>"
date: "<center>2025-10-08</center>"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r Assingment 4}
library(ISLR)
library(dplyr)
library(ggplot2)
library(leaps)
library(lars)


set.seed(123)

#building a second order formula, mains + all pairwise interactions + all squares
second_order_formula <- function(y, xvars)
{
  squares <- paste0("I(", xvars, "^2)", collapse = " + ")
  mains_pairs <- paste0("(", paste(xvars, collapse = " + ") , ")^2")
  as.formula(paste(y , "~" , mains_pairs , "+" , squares))
}

#computing press using hat matrix identity
press_value <- function(fit)
{
  e <- residuals(fit)
  h <- hatvalues(fit)
  sum((e / (1 - h))^2)
}

#aligning names and predicting from the leapsselected lm on a test model matrix
safe_predict_from_leaps <- function(leaps_fit, mm_test)
{
  coefs <- coef(leaps_fit)
  terms <- names(coefs)
  n <- nrow(mm_test)
  Xnew <- matrix(0, n, length(coefs))
  colnames(Xnew) <- terms
  if ("(Intercept)" %in% terms) Xnew[, "(Intercept)"] <- 1
  test_names <- make.names(colnames(mm_test))
  for (nm in terms)
  {
    if (nm == "(Intercept)") next
    pos <- match(make.names(nm), test_names)
    if (!is.na(pos)) Xnew[, nm] <- mm_test[, pos]
  }
  as.numeric(Xnew %*% coefs)
}

#summarizing prediction performance
perf_summary <- function(y_true, y_pred)
{
  r <- suppressWarnings(cor(y_true , y_pred , use = "complete.obs"))
  rmse <- sqrt(mean((y_true - y_pred)^2))
  mae <- mean(abs(y_true - y_pred))
  sst <- sum((y_true - mean(y_true))^2)
  sse <- sum((y_true - y_pred)^2)
  r2 <- 1 - sse / sst
  data.frame(Correlation = r , RMSE = rmse , MAE = mae , R2 = r2)
}


#preparing origin-specific data, 1=usa, 2=europe, 3=japan
prepare_origin_df <- function(origin_val)
{
  df <- as.data.frame(ISLR::Auto) |> dplyr::filter(origin == origin_val) |> dplyr::select(-origin, -name)
  y <- "mpg"
  xvars <- setdiff(names(df) , y)
  xvars <- xvars[sapply(df[, xvars , drop = FALSE] , is.numeric)]
  list(df = df , y = y , xvars = xvars)
}


#fitting leaps and lars and returning predictions and metrics
fit_models <- function(df, y, xvars, train_frac = 0.7, seed = 123)
{
  set.seed(seed)

  #preparing data
  df_use <- df[, c(y , xvars) , drop = FALSE] |> na.omit()

  #splitting into train and test
  n <- nrow(df_use)
  train_idx <- sample.int(n , size = floor(train_frac * n))
  train_df <- df_use[train_idx , , drop = FALSE]
  test_df  <- df_use[-train_idx , , drop = FALSE]

  #building second order matrices
  f2 <- second_order_formula(y , xvars)
  mm_train <- model.matrix(f2 , train_df)
  mm_test  <- model.matrix(f2 , test_df)
  y_train  <- model.response(model.frame(f2 , train_df))
  y_test   <- model.response(model.frame(f2 , test_df))

  #selecting with leaps, first top 5 by cp then picking best by press
  regfit <- regsubsets(mm_train[, -1 , drop = FALSE] , y_train , nbest = 1 , nvmax = ncol(mm_train) - 1 , method = "exhaustive")
  regsum <- summary(regfit)
  top5 <- order(regsum$cp)[1:min(5 , length(regsum$cp))]

  best_press <- Inf
  best_fit <- NULL
  train_cols <- colnames(mm_train)[-1]

  for (i in top5)
  {
    keep <- regsum$which[i , -1]
    chosen <- train_cols[keep]
    fit <- lm(y_train ~ . , data = as.data.frame(mm_train[ , chosen , drop = FALSE]))
    pr <- press_value(fit)
    if (pr < best_press)
    {
      best_press <- pr
      best_fit <- fit
    }
  }

  #predicting with leaps on test
  pred_leaps <- safe_predict_from_leaps(best_fit , mm_test)

  #fitting lars and choosing step by min cp
  x_train <- mm_train[ , -1 , drop = FALSE]
  x_test  <- mm_test[ , -1 , drop = FALSE]
  lars_fit <- lars(x_train , y_train , type = "lar" , normalize = TRUE , intercept = TRUE)
  best_step <- which.min(lars_fit$Cp)
  pred_lars <- as.numeric(predict(lars_fit , x_test , s = best_step , type = "fit" , mode = "step")$fit)

  #metric
  perf <- dplyr::bind_rows(dplyr::mutate(perf_summary(y_test , pred_leaps) , Method = "LEAPS") , dplyr::mutate(perf_summary(y_test , pred_lars) ,  Method = "LARS")) |> dplyr::relocate(Method)

  list(y_test = y_test , pred_leaps = pred_leaps , pred_lars  = pred_lars , perf = perf , leaps_terms = names(coef(best_fit)) , lars_nonzero = coef(lars_fit , s = best_step , mode = "step")[abs(coef(lars_fit , s = best_step , mode = "step")) > 0]
  )
}


#countries analysis

#USA

usa_prep <- prepare_origin_df(1)
usa_res  <- fit_models(usa_prep$df , usa_prep$y , usa_prep$xvars)

ggplot(data.frame(Actual = usa_res$y_test , Predicted = usa_res$pred_leaps) , aes(Actual, Predicted)) + geom_point(color = "#1f78b4" , size = 3 , alpha = 0.85) + geom_abline(slope = 1 , intercept = 0 , color = "#222222" , linetype = "dashed") + labs(title = paste("USA: LEAPS predictions (corr =" , round(cor(usa_res$y_test , usa_res$pred_leaps) , 3) , ")") , x = "actual mpg" , y = "predicted mpg") + theme_minimal()

ggplot(data.frame(Actual = usa_res$y_test , Predicted = usa_res$pred_lars) , aes(Actual , Predicted)) + geom_point(color = "#e31a1c" , size = 3,  alpha = 0.85) + geom_abline(slope = 1 , intercept = 0 , color = "#222222" , linetype = "dashed") + labs(title = paste("USA: LARS predictions (corr =" , round(cor(usa_res$y_test , usa_res$pred_lars) , 3), ")") , x = "actual mpg" , y = "predicted mpg") + theme_minimal()

ggplot(data.frame(Pred = usa_res$pred_leaps , Resid = usa_res$y_test - usa_res$pred_leaps) , aes(Pred, Resid)) + geom_point(color = "#33a02c" , size = 3 , alpha = 0.85) + geom_hline(yintercept = 0 , linetype = "dashed" , color = "#222222") + labs(title = "USA: LEAPS residuals" , x = "predicted mpg" , y = "residuals") + theme_minimal()

ggplot(data.frame(Pred = usa_res$pred_lars , Resid = usa_res$y_test - usa_res$pred_lars) , aes(Pred, Resid)) + geom_point(color = "#ff7f00" , size = 3 , alpha = 0.85) + geom_hline(yintercept = 0 , linetype = "dashed" , color = "#222222") + labs(title = "USA: LARS residuals" , x = "predicted mpg" , y = "residuals") + theme_minimal()



cat("USA performance:")
print(usa_res$perf)

cat("USA: leaps terms:")
print(usa_res$leaps_terms)

cat("USA: lars non-zero coefficients:")
print(round(usa_res$lars_nonzero , 5))




#Europe

eu_prep <- prepare_origin_df(2)
eu_res  <- fit_models(eu_prep$df , eu_prep$y , eu_prep$xvars)

ggplot(data.frame(Actual = eu_res$y_test, Predicted = eu_res$pred_leaps), aes(Actual , Predicted)) + geom_point(color = "#1f78b4" , size = 3 , alpha = 0.85) + geom_abline(slope = 1 , intercept = 0 , color = "#222222" , linetype = "dashed") + labs(title = paste("Europe: LEAPS predictions (corr =" , round(cor(eu_res$y_test , eu_res$pred_leaps) , 3) , ")") , x = "actual mpg" , y = "predicted mpg") + theme_minimal()

ggplot(data.frame(Actual = eu_res$y_test , Predicted = eu_res$pred_lars) , aes(Actual , Predicted)) + geom_point(color = "#e31a1c" , size = 3 , alpha = 0.85) + geom_abline(slope = 1 , intercept = 0 , color = "#222222" , linetype = "dashed") + labs(title = paste("Europe: LARS predictions (corr =" , round(cor(eu_res$y_test , eu_res$pred_lars) , 3), ")") , x = "actual mpg" , y = "predicted mpg") + theme_minimal()

ggplot(data.frame(Pred = eu_res$pred_leaps , Resid = eu_res$y_test - eu_res$pred_leaps) , aes(Pred , Resid)) + geom_point(color = "#33a02c" , size = 3 , alpha = 0.85) + geom_hline(yintercept = 0 , linetype = "dashed" , color = "#222222") + labs(title = "Europe: LEAPS residuals" , x = "predicted mpg" , y = "residuals") + theme_minimal()

ggplot(data.frame(Pred = eu_res$pred_lars , Resid = eu_res$y_test - eu_res$pred_lars) , aes(Pred, Resid)) + geom_point(color = "#ff7f00" , size = 3 , alpha = 0.85) + geom_hline(yintercept = 0 , linetype = "dashed" , color = "#222222") + labs(title = "Europe: LARS residuals" , x = "predicted mpg", y = "residuals") + theme_minimal()


cat("Europe performance:")
print(eu_res$perf)

cat("Europe: leaps terms:")
print(eu_res$leaps_terms)

cat("Europe:lars non-zero coefficients:")
print(round(eu_res$lars_nonzero , 5))



#Japan

jp_prep <- prepare_origin_df(3)
jp_res  <- fit_models(jp_prep$df , jp_prep$y , jp_prep$xvars)

ggplot(data.frame(Actual = jp_res$y_test , Predicted = jp_res$pred_leaps) ,aes(Actual, Predicted)) + geom_point(color = "#1f78b4" , size = 3 , alpha = 0.85) + geom_abline(slope = 1 , intercept = 0 , color = "#222222" , linetype = "dashed") + labs(title = paste("Japan: LEAPS predictions (corr =" , round(cor(jp_res$y_test , jp_res$pred_leaps) , 3) , ")") , x = "actual mpg" , y = "predicted mpg") + theme_minimal()

ggplot(data.frame(Actual = jp_res$y_test , Predicted = jp_res$pred_lars) , aes(Actual, Predicted)) + geom_point(color = "#e31a1c" , size = 3 , alpha = 0.85) + geom_abline(slope = 1 , intercept = 0 , color = "#222222" , linetype = "dashed") + labs(title = paste("Japan: LARS predictions (corr =" , round(cor(jp_res$y_test , jp_res$pred_lars) , 3), ")") , x = "actual mpg" , y = "predicted mpg") + theme_minimal()

ggplot(data.frame(Pred = jp_res$pred_leaps , Resid = jp_res$y_test - jp_res$pred_leaps) , aes(Pred , Resid)) + geom_point(color = "#33a02c" , size = 3 , alpha = 0.85) + geom_hline(yintercept = 0 , linetype = "dashed" , color = "#222222") + labs(title = "Japan: LEAPS residuals" , x = "predicted mpg" , y = "residuals") + theme_minimal()

ggplot(data.frame(Pred = jp_res$pred_lars , Resid = jp_res$y_test - jp_res$pred_lars) , aes(Pred, Resid)) + geom_point(color = "#ff7f00" , size = 3 , alpha = 0.85) + geom_hline(yintercept = 0 , linetype = "dashed" , color = "#222222") + labs(title = "Japan: LARS residuals" , x = "predicted mpg" , y = "residuals") + theme_minimal()


cat("Japan performance:")
print(jp_res$perf)

cat("Japan — leaps terms:")
print(jp_res$leaps_terms)

cat("Japan — lars non-zero coefficients:")
print(round(jp_res$lars_nonzero , 5))

```

#### Overall, LARS was more effective and more consistent than LEAPS in predicting mpg for all three regions, and this provides useful insight into differences in vehicle design and specification priorities. In the U.S., LARS correlated very well with approximately 0.963, had an RMSE of 1.77, and an R2-statistic value of 0.925 (or 92.5%), which indicates there are variables like number of cylinders, vehicle weight, and acceleration that are closely linked to fuel efficiency. Meanwhile, LEAPS did not correlate very well at all - the correlation was near zero (or extremely low) and was accompanied by an enormous error value, or RMSE, indicating these variables from LEAPS did not generalize to new data. This was similarly true in Europe, where LARS again correlated approximately correctly at 0.941, RMSE 2.44, R2=0.876, highlighting the considerable role that displacement, horsepower, and vehicle year play in deciding on mpg. LEAPS had determined, for example, that horsepower did not have much value in prediction, but the accuracy of this prediction can be called into question given what LARS was able to conclude. Again, LEAPS was highly inaccurate (negative correlation, huge error) for every inquiry, which suggested an extreme level of overfitting

#### Japanese vehicles created a more complicated set of circumstances. Even though LARS continued to do better than LEAPS the predictive accuracy for Japanese vehicles was considerably lower than in the other two regions, with a correlation around 0.545, an RMSE of 5.47 and R^2 of 0.217. This suggests that fuel economy for Japanese vehicles is based on a more complicated combination of factors, including interactions between engine power, vehicle weight, and design characteristics. These results reveal clear regional engineering priorities: American manufacturers are focused on maximizing performance while improving fuel economy, European designs are not concerned with the cost performance and mainly maximize technology and engine performance, while Japanese manufacturers blend multiple interacting design decisions in order to increase efficiency.

